Epistasis supports viability under extensive gene-dose insufficiency following chromosome loss

Version 1

Abstract

Abstract

Chromosome loss results in halving the dose of multiple genes. We reasoned that yeast monosomic strains would constitute a relatively simple model suited to address the complex question of how extant diploid organisms can stay relatively fit despite harboring numerous function-canceling (mostly heterozygous) mutations. We started by re-examining diploid yeast strains with a single heterozygous gene deletion and ascertained that many of them produced small but measurable growth defects. Mapped to individual chromosomes, they often combined into burdens sufficient to turn the growth rate negative, that is, inflict lethality. However, the subsequently derived monosomics did experience such loads yet continued to proliferate as if much (often most) of the harm introduced by single mutations disappeared. This constitutes an outstanding example of positive epistasis for fitness. We then sought its functional explanation by analyzing transcriptomes. There was no evidence for widespread gene-dose compensation or cellular stress response. Alterations were abundant but not parallel. A notable exception was the general upregulation of genes coding for ribosomal proteins and the concomitant downregulation of those coding for the proteasome. It indicates that the (irreparably) distorted stoichiometry of ribosomal proteins was the most common and critical impediment to growth and eclipsed the impact of other metabolic insufficiencies. In general terms, the modular structure of the cell leads to effective fragmentation of the total burden of mutations. Those located outside the module(s) currently defining fitness lose at least some of their relevance which produces the epiphenomenon of positive epistasis between individually negative effects.

Introduction

A typical human germline cell has as many as one hundred inactive genes, usually complemented by active alleles (MacArthur et al., 2012). Somatic cell lines in healthy tissues add dozens of new (nearly always heterozygous) mutations every year (Moore et al., 2021). Understanding how this burden of DNA damage translates into the mutational load, defined as a decrease in fitness, has proved to be demanding, not only in the case of humans but also simpler organisms. Estimates of selection and dominance coefficients of single mutations tend to be imprecise, any quantification of interactions between multiple mutations (epistasis) appears even more challenging (Henn et al., 2015). The difficulty with the gene interactions is not just technical but primarily methodological. Any comprehensive research on epistasis would require that the studied combinations of mutations represented an unbiased sample of all possible. The actual combinations encountered in the extant organisms (cells) do not satisfy this requirement because they represent samples distorted by natural selection. For example, the discovery that the individuals whose fitness is negatively impacted by epistasis are underrepresented or even absent cannot be unequivocally interpreted. Either the negative interactions are originally rare or organisms affected by them are eliminated before being scored. Similarly, the presence of multiple and undoubtedly deleterious mutations in an otherwise well-performing organism may mean that either their negative effects tend to alleviate each other or that some yet-unrecognized genetic compensations have developed.

As the fitness architecture of naturally occurring organisms is difficult to disentangle, researchers turned to simpler systems. In the model organism of Saccharomyces cerevisiae, collections of engineered single-gene mutations, such as complete deletions, have been created and used to estimate the selection and dominance coefficients as well as the effects of interaction between a few (typically two) loci (Costanzo et al., 2010; Giaever et al., 2002; Steinmetz et al., 2002). It would be much more difficult to combine large numbers of harmful mutations. This unavoidably long process would expose the strains under construction to a high probability of developing fitness-restoring genetic adjustments which are known to arise even under limited propagation (Szamecz et al., 2014). A rapid way to generate genomes with extensive loads of known mutations is needed. With this in mind, we noted that the budding yeast genome is composed of roughly five thousand genes that are split among 16 chromosomes in a mostly random fashion. We reasoned that monosomy, the loss of one chromosome in an otherwise regular diploid, 2n-1, would mean sudden mutation (partial inactivation) of numerous and arbitrarily assembled genes.

Results

Viability of monosomic strains

We began our study by inserting three markers into one chromosome within each homologous pair. These were two URA3 genes, each at the center of every arm, and one copy of kan close to a centromere (Tutaj et al., 2022). As its homolog was unmarked, mutants lacking the markers signaled a possible loss of the marked chromosome (Materials and Methods). We expected that such generated genuine monosomic strains (M1, M2, …, M16) would exhibit visible growth defects which we indeed observed, except for M1 (Fig. 1a). The loss of markers and slow growth were only hints and proved false on several occasions when sequencing of DNA revealed that the expected loss of a single chromosome did not happen, or it did but was accompanied by other changes. Fig. 1b presents isolates that passed the sequencing test. As Fig. 1c shows, small colonies formed by monosomic cells were regularly infested by much bigger ones. For every monosomic, at least dozens of the latter were re-streaked and found to reproduce the fast-growth phenotype nearly universally. To prevent them from overgrowing the monosomics, we had to keep the cultures small, down to 5 µl of the YPD medium in the case of M5 or M15. The cultures had to be highly replicated so that a sufficient number of those dominated by the slow-growing cells could be still found and recurrently replenished.

Diploid yeast strains lacking single chromosomes (monosomics).

(a) Representative colonies of wild-type diploid (WT) and monosomic mutant strains (M1…M16) when grown together for 24 hours at 30 °C on agar-solidified YPD medium. (b) DNA whole genome sequencing coverage. (c) Examples of large colonies emerging among the small monosomic ones.

The described above procedure could not be completed for three chromosomes—VII, XII, and XIII—even though we paid especially much attention to them. Regarding screens for M7 and M13, we did not find a single mutant that lacked the markers, grew slowly and could rapidly revert to a normal phenotype. Instead, we regularly encountered isolates of the searched genotype without a growth defect and with a diploid genome. Hypothetically, two mitotic cross-overs within a pair of homologous chromosomes could produce one progeny cell without the markers. We were able to reject this possibility based on the results of the present study and a former one (Tutaj et al., 2022). Relevant data and calculations are provided in Supplementary Text. The only viable explanation is that the marked chromosomes VII and XIII had been lost and their non-marked homologs underwent endoreduplication shortly after that. Thus, M7 and M13 were probably not inviable but prohibitively unstable. (They could suffer from very slow growth, a high rate of endoreduplication, or both.) Searching for M12, we obtained unusually numerous and genotypically variable isolates but none correct in terms of growth phenotype or DNA copy number mapping (Materials and Methods, Supplementary Text). Chromosome XII contains the only yeast rDNA region and is a strong recombinational hotspot (Tutaj et al., 2022). Therefore, we suspended attempts to derive M12, especially since, as we show below, its predicted burden of haploinsufficiency was well within the range covered by other chromosomes.

We sought to include all monosomic in our assays in order to test whether monosomy can occur and persist in absence of other genetic changes. In the wild, monosomic isolates have been found only for a few and the shortest chromosomes (Peter et al., 2018). Similar scope of monosomy has been detected in mutation-accumulation experiments (Sui et al., 2020; Zhu et al., 2014). It is unsure whether these monosomies were simply tolerated, compensated by other changes, or conditionally adaptive, especially in natural environments. There have been laboratory experiments signaling that all monosomies can be induced in yeast but this likely happened only transiently as these studies relied on screening for the loss of heterozygosity at selected loci only (Beach et al., 2017; Reid et al., 2008). Thus, former work on yeast monosomy should be taken cautiously and the future one planned properly. Colonies of presumed monosomics can be actually diploids of complex aneuploids. Those initially truly monosomic are readily overtaken by endo-reduplications or other compensations of the chromosome loss (within even a moderately sized colony). Here we combined the marker-based approach with genome-wide tests based on sequencing DNA (reported above) and RNA (below) to confirm that pure, genetically non-compensated, yeast monosomics are viable.

Epistasis for fitness

Before studying epistasis among multiple mutations, their single effects should be known sufficiently well. Regarding the yeast growth rate, effectively none of single gene deletions can make it faster (Sliwa and Korona, 2005). About 1/5 of them stop growth altogether and another 1/5 decelerate it detectably (Giaever et al., 2002). These estimates refer to homozygous or haploid strains under optimal laboratory conditions. Heterozygous deletions, again under good conditions, exhibit morphological alterations quite often but their growth rate, i.e., proliferative fitness, is affected rarely and only modestly (Deutschbauer et al., 2005; Marek and Korona, 2016; Ohnuki and Ohya, 2018). In Materials and Methods, we explain how we selected 468 heterozygous single deletion strains as those likely (haplo)insufficient for growth on rich medium, YPD. We applied replicated growth rate measurements for individual deletion strains and compared them to a set of carefully selected control strains in order to make our estimates possibly precise and reliable. Doubling rates of individual heterozygous deletions were divided by that of the control to yield relative doubling rates, rDRs (Datasheet 1) The frequency distribution of rDR is shown in Fig. 2a. It demonstrates that the negative effects of deletions were undoubtedly present although predictably small, a sizable fraction of them was on the order of the bidirectional effects of the phenotypic plasticity and/or measurement errors (as exposed by the variation within genetically homogeneous control).

Contribution of positive epistasis to fitness of monosomics.

(a) Growth performance of heterozygous single-gene deletion strains tested in this study. Left: list of chromosomes with numbers of assayed deletion strains. Right: frequency distribution of rDR (doubling rate related to that of the control). (b) Growth performance of monosomics. Colored bars represent expected performance calculated as a sum of the single-gene effects per chromosome, rDRE=1+∑di, where di=rDRi−1 (deviation from the control). Gray bars show the observed performance of monosomic strains, rDRM. White arrows mark the expected departure from wild-type fitness (expected genetic load), gray the observed one. Black arrows show the extent and direction of epistasis. Three monosomics were not included in these assays (see the main text).

Epistasis is absent when an effect of a gene variant does not depend on the genetic content of other loci. In the case of fitness, a non-neutral mutation corresponds to a proportional change of wild-type fitness (Crow and Kimura, 1971). An absence of epistasis means that the fitness quotient of a particular mutation remains unchanged irrespective of whether it is the only one present in an individual or accompanied by other fitness-affecting mutations. Accepting this postulate is equivalent to adopting a multiplicative model of fitness structure under which the fitness of a genotype involving n loci is the product of n respective quotients. The model is additive for log-fitness because such a transformation turns a quotient into a deviation from one and a product of quotients into a sum of such deviations. Since we operate with a (log2) transformed measure of fitness, the deviation caused by an i-th gene deletion is equal to di = rDRi−1. The expected effect of multiple loci is the sum of all deviations involved, rDRE=1+∑di. To predict how each of the 16 monosomic strains should perform, we summed deviations caused by the individual deletions present on each chromosome (acquiring 16 values of rDRE). We summed both negative and positive values of d to account for the mentioned above non-genetic variation of estimates. We then compared the predicted values of rDRE with the respective values of rDRM, that is, estimates obtained experimentally for the actual monosomic strains (Datasheet 1). Fig. 2b shows the expected and obtained values of rDR, epistasis is equal to a difference between them. Epistasis between multiple deleterious gene deletions turned out to be not only positive but also large, much larger than that observed for just pairs of deletions (Jasnos and Korona, 2007). Most strikingly, some monosomic strains were expected to have a negative doubling rate, i.e. were predicted to be effectively lethal, but turned out to be able to proliferate. For several monosomics, and especially M4, most of the predicted mutational load was canceled by epistasis. The contribution of epistasis to fitness appeared so high that it might be difficult to accept without functional rationalization.

Shifts in transcriptome profiles

Although epistasis can be viewed in a purely abstract way, a deviation from additivity/multiplicity, it must always have its biological explanation. Our attempt to find it started with the isolation and quantitative examination of mRNA from the ancestral diploid strain and eight randomly chosen monosomic strains. The first question was obvious: was the deficiency in dose compensated by elevated expression? Fig. 3a shows that there was no detectable increase in the intensity of transcription on the monosomic chromosomes. The linear relation between the number of copies of a chromosome and the number of produced transcripts has been reported for yeast, though most of the available data pertain to aneuploids with excessive rather than lacking chromosomes (Larrimore et al., 2020; Torres et al., 2007). Hence, neither our study nor earlier ones revealed compensation at the level of chromosome-scale averages. Those averages are dependent on hundreds of genes, mostly dose-insensitive, and therefore they would be insensitive to an increased expression of the genes which were important but relatively few. In this experiment, we knew for which genes upregulation would be most rewarding and could focus our attention on them. Fig. 3b shows that they were not transcribed at elevated rates. Thus, neither physical underrepresentation (single copy) nor functional importance (impact on growth) triggered transcriptomic compensation.

Absence of transcriptional compensation in monosomic strains.

(a) Halved RNA production on monosomic chromosomes. For every ORF, the obtained number of RNA-seq reads was divided by the number expected for it under expression being constant over an entire genome. (b) Expression under monosomy vs. single-deletion fitness (rDR). X-axis: the effect of monosomy on a particular mRNA level was calculated as log2 fold change relative to the diploid control and displayed as a color-coded bar at its chromosomal position (the width of a box reflects chromosome length). The shifts in expression at the monosomic chromosomes were calculated after taking into account that the gene dose was halved there. Y-axis: the difference in rDR between a single gene deletion strain and the control, d=rDR-1. Pairs of estimates were tested for rank correlation; shown are Spearman’s coefficients and p-values (not corrected for the multiplicity of comparisons).

The availability of entire transcriptomes enables examination of the cell functioning far beyond testing for compensatory expression at the affected loci. We aimed at finding any distinct and functionally interpretable patterns of adjustments that could explain why the monosomics performed better than expected. We found multiple statistically significant up- and downregulations in transcriptomes of every analyzed monosomic strain (Datasheet 2). Both parallelisms and incongruencies between them were visible. Fig. 4a shows the similarities.

Parallel and divergent shifts in transcriptomes of monosomic strains.

Heat maps show monosomic mRNA frequencies divided by respective diploid (control) ones. (a) Gene Ontology categories selected to demonstrate similarities in transcriptional profiles of monosomic strains. (b) Regulons demonstrating differences in gene expression between monosomic strains. Expanded versions of all panels can be found in Supplementary Figure S1.

The translational apparatus of the cytosol was on average considerably and universally upregulated, as evidenced by transcripts coding for proteins building both large (e.g. RPL28) and small (RPS29B) ribosomal subunits. Cytosolic proteolysis was downregulated (core proteasome constituent PRE6). In a remarkable accord, chaperones needed to fold newly synthesized peptides were upregulated (SSB1/2) while those required to direct destabilized chains to degradation were downregulated (SSA1, HSP82, HSP104). Transcripts coding for mitochondrial proteins were generally less abundant, with a marked decline in the expression of several genes coding for the electron transport chain (Datasheet 2). At the same time, the expression of antioxidative machinery was dampened, especially that of a chief ROS scavenger (SOD2).

To find transcriptomic differences, we examined groups of genes each reacting to a known signal. This raised chances for detecting statistical significance and functional divergence. In our experiment, all strains received identical outer signals. Therefore, different expression patterns would indicate different internal distortions. Indeed, as Fig. 4b demonstrates, substantial variation was detected within several regulons: cAMP-PKA (glucose-activated signaling), NCR (nitrogen catabolism repression), GAAC (general amino acids control), and Msn2up (activation upon a broad array of stress factors).

Discussion

The successful derivation of monosomic strains demonstrated that it is possible to obtain an a priori designed set of strains with hundreds of partially inactive genes with tens of them exerting deleterious fitness effects. It implies that the ability to remain viable despite carrying extensive mutational burdens was not restricted to assorted combinations of affected genes but would likely hold for a broad variety of them. The viability was supported by the epistatic effect which was of positive value and remarkable strength. The contribution of epistasis was probably still underestimated. It has been recently exposed that many of the single deletions with negative fitness effects had undergone compensatory evolution under recurrent propagation of the strains carrying them (Puddu et al., 2019). Therefore, some of those deletions were likely missing among our strains or were partially compensated which made their joint predicted harm lower than it would be in an unaffected collection. All those potentially missing heterozygous effects were present in the freshly derived monosomic strains biasing downward the difference between obtained and expected fitness, that is, the estimated contribution of epistasis. This caveat actually strengthens our claim of detecting the positive epistatic component that is sizable and can even offset most of the total harm associated with single negative effects.

The ascent of systems biology has raised hopes that the “statistical” and “biological” sides of epistasis can be merged coherently (Moore and Williams, 2005; Phillips, 2008). Indeed, a truly functional explanation of the measured growth effects would require understanding how the material and energetic expenditures were altered by the same mutations when they were separate or combined. Such an ultimately metabolomic interpretation could be attempted only semi-quantitatively and for much simpler systems (Molenaar et al., 2009). Regarding proteome and transcriptome, the latter has the advantage of higher accuracy and repeatability, as well as sensitivity to products of weakly expressed genes. The relation between transcriptome and proteome is generally rigid in yeast. In yeast aneuploids, the bulk of changes observed in mRNA are generally reflected as proportional shifts in both the profile of translation and mature protein composition (Larrimore et al., 2020; Pavelka et al., 2010). The relation is not perfect as, for example, disomics do have a somewhat attenuated average expression of proteins coded at the doubled chromosome, but the outcome is still much closer to doubling than parity (Dephoure et al., 2014). The post-transcriptional attenuation can be more frequent and pronounced when single genes are doubled but this is a case much different from ours (Ascencio et al., 2021). Most importantly, the yeast transcriptome is currently unparalleled in providing broad and comprehensible insights into cell signaling and functioning under major transitions such as fast-slow growth, anabolism-catabolism, presence-absence of stress. In sum, although the analysis of mRNA obviously does not provide a full description of the cell’s state, it is a sensible introduction to it.

The simultaneous upregulation of genes coding for ribosomal proteins (RP) and downregulation of those coding for subunits of the proteasome was observed in all monosomic strains and is likely providing crucial insight into the functional reaction of the cell to monosomy. An absence of a single chromosome meant halving the dose of several genes coding for the translational apparatus that resided on it while tens of others remained unaffected. The cell could not restore the required stoichiometry by targeted overexpression of the affected genes. The favorable environmental conditions signaled that translation had to be intensified but the functioning ribosomes were scarce resulting in apparently indiscriminate overproduction and withholding of degradation of all needed proteins. The reaction was inadequate and costly, but it should be seen as an attempt at ad hoc rebalancing, not a prepared (evolved) response. This interpretation, strongly suggested by the transcriptomic data, appears still more plausible after considering the functional diversity of the 468 genes selected as haploinsufficient. Table 1 shows that the diversity was ample: of the one hundred Slim GO Biological Process categories, dozens could be linked to every chromosome-associated subset of deletions. The number of affected processes was higher than the number of deletions because a single deletion often belonged to several categories (Saccharomyces Genome Database). However, there was a telling common motif. The processes involved in the “protein synthesis apparatus” (our designation) were most often represented. Their predicted harm was also especially high, typically sufficient or exceeding that required to reduce the growth rate to the levels actually observed in monosomics. The distortion of biosynthesis was thus real and its relation to the observed transcriptomic reaction was probably causal, not merely coincidental.

Some of the other adjustments seen in the monosomic transcriptomes could be also linked to the dominant effect of translational inefficiency. A signal to increase biosynthesis would mobilize fermentation and curb oxidation which downregulated genes coding for the ROS-scavenging proteins. The Mns2-mediated and other stress responses were seemingly absent, contrary to previous reports (Sheltzer et al., 2012). But, among the hallmarks of stress reactions are decreased RP production and increased proteolysis (to remove destabilized proteins). These two processes were here pushed in opposite directions by the loss of RPs stoichiometry blurring the expected patterns. The non-homogeneity of reaction in other analyzed regulons—cAMP-PKA, NCR, GAAC—means that functions other than ribosome assemblage were affected seriously but differently, depending on the composition of genes made insufficient by the loss of a chromosome. The environment was the same and rich, with nothing in short supply, therefore perceived deficiency of metabolites resulted from internal distortions.

Datasheet 2 shows a very large number of differences between monosomics in the expression of single genes, the variation that would be difficult to summarize or, still more, to fully understand in functional terms. If only some of the challenges met by individual monosomics were similar, while so many others disparate, then why was the uniformly positive epistasis observed?

A possible answer starts with the realization that the cell is an aggregate of multiple functional modules (Hartwell et al., 1999). Our monosomic strains had many different modules (recall the Slim GO categories) that were negatively affected by partially inactivating mutations. These mutations did not interact with each other directly but rather through the modules they belonged to. Mutations that affected the most critical module(s) suppressed the negative effect of those mutations that impaired other, non-limiting, modules. The latter were either less damaged or less needed under current conditions. The epiphenomenon of positive epistasis for fitness reflected the fact that not all mutations exerted their negative effects, at least not at a full scale. In our case, it is not just speculation but rather the simplest and most prudent explanation linking the growth rate and transcriptomic data. Our experimental system is in a way special: a microbial cell with multiple modules compromised but exposed to favorable conditions and thus tested for the single capacity of growing fast. However, it is also illuminating. A unicellular organism is typically tested for only a few fitness-defining functions, they would be much different under prolonged starvation. The specialized cells of a complex organism are required to uphold only some, and different, elements of life. The modularity of the cell helps to understand how so many partial genetic insufficiencies can be carried by so many different organisms without causing a drastic decline in their fitness.

Interactions within pairs of gene knock-outs have attracted much attention and some attempts to understand the uncovered variety of fitness effects have relied on the concept of cellular modularity (Costanzo et al., 2020; Segrè et al., 2005; Szappanos et al., 2011). High-order epistasis for fitness, arising from interactions between at least several mutations, has been often mentioned in the context of “diminished returns” of consecutive beneficial variants (Barrick et al., 2009; Chou et al., 2011; Khan et al., 2011; Kryazhimskiy et al., 2014). It has been proposed that the pattern emerges from interactions of mutations through some ‘global’ phenotypic interface, e.g., fitness (MacLean et al., 2010; Perfeito et al., 2014; Schoustra et al., 2016). However, a similar overall effect could be theoretically generated by the entirety of ‘idiosyncratic’ interactions, without any intermediating factors (Lyons et al., 2020; Reddy and Desai, 2021). Indeed, this has been demonstrated experimentally, at least for a selected and moderately large set of mutations (Bakerlee et al., 2022). Finally, the diminishing returns epistasis of beneficial mutations has been attributed to cellular modularity by postulating that the contribution of each functional module to fitness must have its upper limit (Wei and Zhang, 2019). Considering that the negative epistasis between multiple positive effects can result from different mechanisms, we do not insist that the proposed by us explanation of positive epistasis between multiple negative effects is the only one possible. However, it can be especially applicable when the deleterious mutations are really numerous and distributed widely, though not necessarily evenly, across many cellular subsystems.

Materials and Methods

Single gene deletion strains

Selection of single gene deletions with a possible effect of haploinsufficiency

Deutschbauer et al. have assayed a complete collection of heterozygous single gene deletion strains and identified a total of 184 genes, 98 essential and 86 non-essential, as haploinsufficient for growth in rich medium, YPD (Deutschbauer et al., 2005). We included this set of genes in the present study. Using a different technique, Marek et al. have tested al1 142 essential heterozygous single-gene deletions and 946 non-essential ones selected as likely non-neutral for growth (Marek and Korona, 2016). We reviewed the latter study and, setting a level of false discovery rate at 0.15, accepted as many as 404 genes, 256 essential and 148 non-essential, as potentially haploinsufficient. The two sets obtained in two different studies, 184 and 404, overlapped in 112 cases. The overlap was clearly higher than expected, 14.3, if the two sets were just random samples from among 5,200 yeast genes. On the other hand, it was limited and suggested new growth assays were desirable. The new assays are described below, they involved all unique strains identified in the two studies minus 5 which were not present in our present strain collections and another 3 which were dropped as superfluous for the described below orthogonal design of experimental blocks. (They were also the 3 least promising based on the former assays.) Datasheet 1 lists the strains.

Control for single deletion strains

To correctly quantify the predictably small negative growth effects introduced by heterozygous single-deletion strains, an unbiased and precise estimate of a wild-type phenotype is needed. In the present study, we sought to accomplish this by setting not a single but multiple strains as a control. The reason was that the used here gene deletions were constructed over several years by multiple laboratories and were then handled with repeated rounds of propagation that could result in genetic divergence. We therefore considered it risky to choose a single strain as a control for all others. We looked for a group of strains that would be indistinguishable in terms of growth rate suggesting that they have acquired genetic alterations affecting this trait. In the case of non-essential genes, we reviewed the Saccharomyces Genome Database and found 25 ORFs termed “dubious”, at present nearly sure to be false ORFs, which were located between other ORFs and no closer to their START or STOP codons than 100 bp. Heterozygous strains carrying deletions of these genes were assayed for doubling rate in a way specific to this study (described below). After replicated assays, 16 deletion strains closest to the medium growth performance, and statistically not different from each other, were selected as the control ones. Knowing that the non-essential and essential strains differed somewhat in their origin and subsequent handling (Brachmann et al., 1998), we decided to derive a separate set of control strains for the latter. We selected 32 essential genes which were in the very center of a single, strong and narrow modal peak of the frequency distribution of estimates collected for the whole collection of heterozygous essentials (Marek and Korona, 2016). Again, after replicated tests of growth rate, 16 strains closest to the median value and not different from each other were selected to constitute the final control for essential gene deletions. Control strains of both groups are listed in Datasheet 1.

Monosomic strains

Parents of monosomic strains

In a former study, we used 32 strains that had one counter-selectable marker (URA3) close to the center of each of the 32 chromosomal arms and a drug resistance marker (kan) close to a centromere (Tutaj et al., 2022). Pairs of strains with the same centromere marker and the counter-selectable markers residing on either the left or right arm of the same chromosome were mated with each other. The resulting diploids were sporulated and tetrads dissected to get triple-marked haploid genotypes, URA3-kan-URA3. The latter were mated with a standard BY haploid strain of the opposite mating type. In the final set of 16 diploid strains, each strain had one chromosome triple-marked while its homolog and the rest of the 15 chromosome pairs were isogenic with the diploid strain BY4743.

Derivation and verification of monosomic strains

The diploid strains with triple-marked chromosomes were grown overnight in synthetic complete medium (SC) and then 50 to 500 µl samples of the resulting cultures were overlaid on standard 5-FOA (5-fluoroorotic acid) plates. Emerging colonies were transferred onto new 5-FOA plates and YPD plates with 200 mg/ml of geneticin. The goal was to identify variants that were able to grow on the 5-FOA but not geneticin plates as this indicated that all three marker genes were possibly lost together with an entire chromosome. Colonies identified in this way were often of different sizes suggesting that they were genetically heterogeneous. The next criterion was to find variants that would form visibly smaller colonies than those of the parental strain but rarely, though regularly, produce colonies similar to the latter. Such strains were grown as replicate small cultures (5 to 100 µl dependent on particular monosomic strain), tested for negligible frequency of cells forming large colonies, and collected into larger samples enabling isolation of DNA at amounts sufficient for the next generation sequencing. Monosomy was considered confirmed when the number of reads was halved for the whole stretch of a single, and the expected one, chromosome. Although simple, this protocol required multiple attempts to be completed for some chromosomes as few colonies tended to appear on the 5-FOA plates, most of them remained resistant to geneticin and those which passed these two criteria did not show the required reversibility, that is, the tendency to return occasionally to normal growth in apparently one step. Even if all these phenotypic criteria were met, the sequencing occasionally uncovered genomes other than those of pure monosomics, that is, with the whole and only single-chromosome removed. Nevertheless, once confirmed, the monosomic strains could have been propagated reliably on rich and synthetic media, including that with 5-FOA, as long as the recurrent emergence of fast-growing colonies was monitored and countered.

Control for monosomic strains

The monosomic strains were all derived by us and the derivation involved our stock of haploid BY4741 and BY4742 strains. These two were then crossed to obtain a diploid BY4743, devoid of the URA3MX4 and kanMX4 cassettes, and then used as a control in growth assays of the monosomic strains.

Estimation of DR and rDR

The collection of single gene deletion strains was arrayed on 6 flat-bottom 8×12 well titration plates with 150 µl aliquots per well. Within each plate, the first and last well contained clean YPD medium, rows 3 and 10 hosted control strains, and deletion strains occupied the rest of the plate. Plates were filled with either essential or non-essential deletions accompanied by the respective 16 control strains mentioned above. One plate contained both essential and non-essential deletions together with 8 essential and 8 non-essential control strains. Plates were inoculated from thawed samples at 1-5% and kept non-agitated for 48 hours at 30 °C when they reached approximately similar densities of stationary phase cells. Such conditioned micro-cultures were used to inoculate plates with fresh YPD at 0.5% and kept at 30 °C with 1,000 rpm agitation. The cultures were tested for OD (600 nm) every 0.5 h with a TECAN Infinity reader. Four independent replicates of the measurements, starting with independent conditionings, were carried out. OD reads were used to calculate DR (doubling rate). To get rDR (relative doubling rate), every DR estimate of an experimental strain was divided by an average DR of control strains residing on the same plate. In the case of the monosomic strains, the whole protocol was analogous except that all experimental and control strains were held within one plate. The cultures of monosomic strains used in this assay were tested to contain less than 1% of fast-growing cells at the time of OD measurements.

Analysis of DNA and RNA

An initial step in preparing for both the DNA and RNA analysis was to collect samples of monosomic cells that would be nearly free of the fast-growing revertant cells. Individual monosomic strains were grown in YPD at 30 °C up to stationary phase as replicate micro-cultures (5 to 100 µl). The latter were serially diluted and plated to test for the appearance of large colonies suggesting the rise of compensatory mutations. The micro-cultures in which colonies typical for a particular monosomic strain constituted more than 99% were pooled together. Such tested stationary cultures of monosomic cells were directly used to extract DNA as a template for high-coverage sequencing (PE 150, expected read depth ∼80). The resulting reads were mapped, using bowtie2 (Langmead and Salzberg, 2012), along standard sequences of yeast chromosomes: Ensembl release 100, S. cerevisiae genome R64-1-1. Duplicated reads were marked with MarkDuplicates (Picard Toolkit 2019. Broad Institute, GitHub). Samtools (1.15.1) were used for BAM files sorting, indexing and coverage analysis (Danecek et al., 2021).

In the case of RNA, the tested for purity cultures of monosomic cells were transferred to fresh YPD and incubated with agitation at 30 °C for 4 hours. Total RNA was then extracted with the RiboPure™ RNA Purification Kit. Three replicas of the monosomic and the control BY4743 strain were prepared in this way. Libraries preparation and PE 150 sequencing was carried out by Novagene. About 20 mln read pairs were obtained per sample. Quality control of the reads was performed with fastQC v0.11.9 (Andrew, 2010). The RNA reads were mapped to the above-specified standard sequence with Hisat2 v2.1.0 (Kim et al., 2015). The resulting alignment files were sorted and indexed with samtools (1.9). Transcript quantification was performed with cuffquant/cuffnorm v2.2.1 (Trapnell et al., 2012). Gene count data normalization (“TMM” method) and differential expression analysis (exact test) were performed in the EdgeR test(Robinson et al., 2010).

Supporting information

eLife Assessment

This study presents important findings that set a new standard for generating and maintaining monosomic yeast lines, and is the first, to our knowledge, to assess gene expression in yeast monosomies. The research offers an innovative approach to evaluate epistasis between genes on the same chromosome; however, the strength of evidence supporting the conclusions on positive epistasis for fitness is still incomplete due to the need for clarification for the exact calculation method of expected fitness without epistasis, as well as due to a lack of molecular mechanisms. Additionally, the interpretation of transcriptome results would benefit from a more comprehensive consideration of previous studies on genomic balance and the inclusion of further analyses and experiments.

Reviewer #1 (Public Review):

The study by Korona and colleagues presents a rigorous experimental strategy for generating and maintaining a nearly complete set of monosomic yeast lines, thereby establishing a new standard for studying monosomes. Their careful approach in generating and handling monosome yeast lines, coupled with their use of high-throughput DNA sequencing and RNA sequencing, addresses concerns related to genomic instability and is commendable. However, I would like to express my concerns regarding the second part of the study, particularly the calculation of epistasis and the conclusion that vast positive epistatic effects have been observed. I believe that the conclusion of positive epistasis for fitness might be premature due to potential errors in estimating the expected fitness.

The method used to calculate fitness expectation (1 + sum(di), where di = rDRi - 1) may be inappropriate. By reading Figure 2a, it appears that the authors defined rDR as log(mutant growth rate)/log(wild-type growth rate), but I am unsure about the biological meaning of 1 + sum(di) here. In other words, what does it exactly mean when a negative y-axis value is observed in Figure 2b if it is a relative doubling rate? I would assume that the log transformation should be performed after (rather than before) dividing the mutant growth rate by the wild-type growth rate (i.e., log(mutant growth rate/wild-type growth rate)). I believe the expected growth rate for a monosome should be calculated as exp(sum(log(mutant growth rate i/wild-type growth rate))), which can then be compared with the wild-type (with a value equal to 1). Based on this calculation method, if gene A exhibits a 20% reduction in fitness when halved (A/-) and gene B exhibits a 30% reduction (B/-), the expected fitness of A/- B/- should be 56%. Therefore, it is unclear how exactly the expected fitness without epistasis was calculated and how that would affect the estimation of the sign and quantity of epistasis.

While widespread positive epistasis in yeast has been reported by other studies (e.g., doi: 10.1038/ng.524, but not to the extent reported in this study), the conclusion of the current study might not be sufficiently supported. I recommend that the authors revisit their calculation methods to provide a more convincing conclusion on the presence of positive epistasis for fitness in their dataset. Overall, I appreciate the authors' efforts in this study, but believe that addressing these concerns is essential for strengthening the validity of their findings.

Reviewer #2 (Public Review):

This study examines most monosomies in yeast in comparison to synthetic lethals resulting from combinations of heterozygous gene deletions that individually have a detrimental effect. The survival of monosomies, albeit with detrimental growth defects, is interpreted as positive epistasis for fitness. Gene expression was examined in monosomies in an attempt to gain insight into why monosomies can survive when multiple heterozygous deletions on the respective chromosome do not. In the RNAseq experiments, many genes were interpreted to be increased in expression and some were interpreted as reduced. Those with the apparent strongest increase were the subunits of the ribosome and those with the apparent strongest decreases were subunits of the proteasome.

The initiation and interpretation of the results were apparently performed in a vacuum of a century of work on genomic balance. Classical work in the flowering plant Datura and in Drosophila found that changes in chromosomal dosage would modulate phenotypes in a dosage sensitive manner (for references see Birchler and Veitia, 2021, Cytogenetics and Genome Research 161: 529-550). In terms of molecular studies, the most common modulation across the genome for monosomies is an upregulation (Guo and Birchler, Science 266: 1999-2002; Shi et al. 2021, The Plant Cell 33: 917-939).

In the present yeast study, not only are there apparent increases for ribosomal subunits but also for many genes in the GAAC pathway, the NCR pathway, and Msn2p. The word "apparent" is used because RNAseq studies can only determine relative changes in gene expression (Loven et al., 2012, Cell 151: 476-482). Because aneuploidy can change the transcriptome size in general (Yang et al., 2021, The Plant Cell 33: 1016-1041), it is possible and maybe probable that this occurs in yeast monosomies as well. If there is an increase in the general transcriptome size, then there might not be much reduction of the proteosome subunits as claimed and the increases might be somewhat less than indicated.

It should be noted that contrary to the claims of the cited paper of Torres et al 2007 (Science 317: 916-924), a reanalysis of the data indicated that yeast disomies have many modulated genes in trans with downregulated genes being more common (Hou et al, 2018, PNAS 115: E11321-E11330). The claim of Torres et al that there are no global modulations in trans is counter to the knowledge that transcription factors are typically dosage sensitive and have multiple targets across the genome. The inverse effect trend is also true of maize disomies (Yang et al., 2021, The Plant Cell 33: 1016-1041), maize trisomies (Shi et al., 2021), Arabidopsis trisomies (Hou et al. 2018) and Drosophila trisomies (Sun et al. 2013, PNAS 110: 7383-7388; Sun et al., 2013, PNAS 110: 16514-16519; Zhang et al., 2021, Scientific Reports 11: 19679; Zhang et al., genes 12: 1606). Taken as a whole it would seem to suggest that there are many inverse relationships of global gene expression with chromosomal dosage in both yeast disomies and monosomies.

To clarify the claims of this study, it would be informative to produce distributions of the various ratios of individual gene expression in monosomy versus diploid as performed by Hou et al. 2018. This will better express the trends of up and down regulation across the genome and whether there are any genes on the varied chromosome that are dosage compensated. The authors claim there are no genes that are compensated on the varied chromosome but considering how many genes are upregulated across the genome, it would seem that a subset are probably upregulated on the cis chromosome as well and approach the diploid level, i.e. are dosage compensated. A second experiment that would clarify the results would be to perform estimates of the general transcriptome size. If the general transcriptome size is actually increased, the claims of reduced expression of the proteosome might need to be revised (See Loven et al., 2012 for an explanation).

Reviewer #3 (Public Review):

The current study examined 13 monosomic yeast strains that lost different individual chromosomes. By comparing the fitness of monosomic strains and several heterozygous deletion strains, the authors observed strong positive epistasis for fitness. The transcriptomes of monosomic strains indicated that general gene-dose compensation is not the reason for fitness gains. On the other hand, gene expression of ribosomal proteins was up-regulated and proteasome subunit expression was down-regulated in all tested monosomic strains. The authors speculated that overexpression in combination with decreased degradation of the insufficient proteins might explain the positive epistasis observed in monosomic strains. This study investigates an important biological question and has some interesting results. However, I have some reservations about the data interpretations listed below.

1. In Figure 3b (and line 179), the authors stated that those haploinsufficient genes were not transcribed at elevated rates, but almost half of them are in reddish colors (indicating that the expression is higher than 1-fold). Obviously, many haploinsufficient genes are up-regulated in monosomic strains. What the data really show is that the level of overexpression is not correlated with the fitness effect of the deletion (since all the p values are not significant). The authors need to correct their conclusions.

2. Why are some monosomic strains removed from the transcriptomics analysis, especially when the chromosome IV and XV strains show very strong positive epistasis? The authors need to provide an explanation here.

3. The authors stated that diploidy observed in chromosome VII and XIII strains were due to endoreplication after losing the marked chromosomes (lines 97 and 117). Isn't chromosome missegregation an equally possible explanation? Since monosomic cells are generated by chromosome missegregation during mitosis, another chromosome missegregation event may occur to rescue the fitness (or viability) of monosomic cells in these strains.

Author response:

The following is the authors’ response to the original reviews.

Public reviews:

Response to Reviewer #1 (Public Review):

The reviewer is correct that the previous explanation of the fitness calculation could be considered insufficient as it was only briefly described in Results. In the revised manuscript, in the "Supplementary Materials" section and then in "Supplementary Text 1", we provide a full definition of the fitness of strains carrying single or multiple mutations and thus show how epistasis was calculated.

Response to Reviewer #2 (Public Review):

In our opinion, the reviewer's comments relate to three issues. First, our finding that the level of transcription of the monosomic chromosomes is not upregulated was not compared with the results of other studies, including those in other organisms. Indeed, we did not mention that the gene dosage distortions introduced by aneuploidy are frequently and profoundly compensated in multicellular organisms. We cite the suggested broad and recent review paper in the revised manuscript (line 247). We also removed the somewhat provocative sentence: “The relationship between transcriptome and proteome is generally fixed in yeast”. Regarding this organism, both data and opinions remain indeed conflicting when considering the work with many different yeast strains. But the standard laboratory strains stand out as those where dosage compensation is absent or weak. A paper published a year ago states flatly: "... at least in the strain background used here (authors: BY, the same we use), aneuploidies are transmitted to transcriptome and proteome with a minimum of gene-dosage buffering, rendering aneuploidies discoverable by proteomics" (Messner et al. 2023). A more recent paper reports: "In lab-generated aneuploids, some proteins - especially subunits of protein complexes - show reduced expression, but the overall protein levels correspond to the aneuploid gene dosage" (Muenzner et al. 2024). This "reduced expression" was seen in disomics and was achieved by upregulated proteolysis, whereas we have monosomics and downregulated proteolysis. In summary, we cannot back away from our claim that the biases introduced by monosomy were not compensated. (It is not critical to our paper, we could do it and still leave our main claim about extraordinarily high positive epistasis intact). Muezner and colleagues do report compensation, but in "wild" strains. Our explanation would be that the existing yeast aneuploids are not a random sample of aneuploid mutations. In particular, they could be strains, perhaps relatively rare, in which the genetic background was permissive for aneuploidy from the start or allowed rapid evolution toward tolerance of aneuploidy. Strains with rigid gene-mRNA-protein relationships suffer so much that they perish unless they are shielded from selection, as is possible in the laboratory. The reviewer will know better whether this might also apply to multicellular organisms.

The second concern is that we did not sufficiently report "... the trends of up- and downregulation across the genome and whether there are any genes on the varied chromosome that are dosage compensated". We believe we have indeed done this, albeit mostly in a simple graphical fashion. For the whole genomes, Datasheet 2 reports the extent of down- or up-regulation for each gene in each strain and highlights those that are statistically significant. We do not analyze the distributions of these deviations because they are relative. They represent individual gene down- and up-regulations within a monosomic transcriptome compared to the corresponding genes in the diploid transcriptome, with the total size of the transcriptomes set equal. Thus, the downs and ups cancel each other out, the left and right sides of the distribution would be equal in their totals, and we have no meaningful expectations about the possible variation in the shapes of the overall distributions or their opposite sides. As for the "varied chromosome", we show that there were extensive down- and up-regulations on the monosomic chromosomes, even though the mean expression for them was half that of the diploid chromosomes. This can be seen in Figure 3B as blue and red colored bars that are present on each monosomic chromosome and intermingled along its length. The purpose of these graphs is to show that even the genes in which the halving of the dose was most damaging to fitness (most negative values of rDR-1) did not tend to be upregulated on average (both blue and red colors are found among them). We consider this an important and original part of our data.

Finally, the reviewer is concerned that we are only dealing with the relative abundance of mRNA species. He/she suggests that "... an experiment that would clarify the results would be to perform estimates of the total transcriptome size. If the general transcriptome size is indeed increased, the claims of reduced proteosome expression may need to be revised". We followed this advice and extracted transcriptomes from known amounts of yeast cells with known amounts of standard mRNA or "spike" added. We thus seriously considered the reviewer's suggestion, even though it was contrary to our intuition and, we believe, was not confirmed in the additional experiment. The results are reported in the last paragraph of Results and shown in Supplementary Figure S3. Our arguments are listed in that paragraph, so we will not repeat them here.

Response to Reviewer #3 (Public Review):

(1) Figure 3b – both its legend and reference to it in the main text are corrected in line with suggestions made by Reviewers #1 and #3.

(2) We had to restrict our mRNA analysis to about a half of strains. We decided for purely random selection. It left M4 outside but nevertheless included M2, M10 or M16 representing the strains with especially high level of epistasis. See msc. lines 161-162.

(3) We agree, and say so in the article, that both the loss and gain of a copy of a chromosome most likely result in errors in mitosis. By "endoreduplication" we mean any event resulting in two chromosomes instead of one, not necessarily additional DNA replication as we now clarify. We also suggest that both loss and endoreduplication occurred in all strains, but in M7 and M13 they happened so close together that we could not isolate the rare monosomic cells from the rapidly spreading revertants (lines 86-91).

Recommendations for the authors:

Reply to Reviewer #1 (Recommendations for The Authors):

The legend to Fig. 3b is hopefully clearer now.

Reply to Reviewer #2 (Recommendations for The Authors):

We understand that these points were raised also in the public review so the answer to the latter is also relevant to the recommendations for authors.

Reply to Reviewer #3 (Recommendations for The Authors):

(1) The first sentence of this comment may be based on a misinterpretation of our main argument. We believe that the upregulation of ribosomal protein (RP) coding genes was not helpful, but harmful. It was costly because RPs are a large part of the proteome, but it did not help translation because it did not restore the stoichiometry of RPs. This unproductive investment reduced the rate of remaining metabolism, so that other impairments introduced by halving the doses of other genes were no longer critical, and this made them unobservable at the level of phenotype, i.e., produced epistasis. However, both this Reviewer and Reviewer 2 seem to suggest that an entire translational apparatus may have been expanded, compensating for its reduced efficiency (per transcript). Reviewer 2 suggested an mRNA spike as a standard, and we followed this approach as more accessible to us. (We reiterate our claim of good agreement between mRNAs and proteins in the BY strain, supported by two new important papers, line 256-257). The results are reported in the last paragraph of Results. We believe that they indicate a reduction, not an increase, in the translational apparatus (including its parts encoded on the monosomal chromosomes), so that our explanation of positive epistasis remains unchallenged.

(2) We re-examined the sequences and found that there were heterozygous SNPs in the same gene, RSP5, in several strains. One was a loss of a START codon (M3, M4, M6, M8, M9, M10, M14, M16), always the same. The other was a substitution, always the same, in M5, M11 and M15. There were no mutations in this gene in M1 and M2. We tested our stock haploid strains BY4741 and 4742 and found that they were not mutated. However, we also recovered the specific haploid strains used in the final crosses to construct the diploid strains used to obtain monosomics. Some had one of the two mutations, some were clean. All grew normally, the mutants were similar to the wild types, indicating that the fitness effect of the mutations, even in haploids, was at most partial, since the expected severe effects of RSP5 inactivation were not visible.

Where do the mutations come from? In previous experiments, we subjected some BY strains to severe selection regimes. As we can now surmise, mutations in RSP5 helped to resist some of them, especially those involving overexpression of selected genes. (We do not summarize here the results of our lengthy review of our notes and the literature leading this explanation to be the most plausible). Unfortunately, we used strains that went through that harsh selection in crosses serving to derive another collection of strains, those used here.

How critical is it? First, the mutations were heterozygous, which further reduced their apparently weak effects. Second, M1 and M2 were free of them. Third, we tried to get clean monosomics, i.e. with type homozygous for RSP5. We obtained such strains with monosomy as the only change for M9, M10 and M16. The other three attempts did not yield correct M3, M5 and M6, but complex aneuploids. This is normal, as we explain (complain) in Results. We would have to isolate a large number of potential monosomies and then sequence them to show that all exact monosomies can be derived in the absence of mutations in RSP5. We believe that after an effort comparable to that required to obtain the first set of monosomics, we would complete it. For financial and organizational reasons, this is not possible at this time. We do not consider it necessary to complete the revision. Note that of the five mutation-free straight monosomics, M2, M10 and M16 are among the most affected and thus have the highest positive epistasis. Yes, the role of point mutations cannot be excluded for other monosomics, although we strongly believe it is unlikely. But we have removed all our previous claims that our monosomies are certainly not supported by other genetic changes. Most importantly, our main claim of positive epistasis in its purely descriptive genetic sense remains unaffected. The main functional argument also holds: the indiscriminate overproduction of unbalanced RP proteins was so costly that inefficiencies introduced in functional modules other than biosynthesis become much less relevant. Thus, the main message of our work does not depend on the thinkable, in our view unlikely, role of mutations in RSP5.

We provided this lengthy explanation to show that we cared about the reviewer's comment and tried to deal with it in an honest way. It was a lot of pain and no gain for us, but we are still grateful for the opportunity to re-examine our main claims.

(3) The 16 (non-essential) plus 16 (essential) strains were replicated 3 times each. In preliminary experiments, we tested that they were not statistically different (using one-way ANOVA). We considered these 32 strains to have the same genetic background, and thus we considered the 96 estimates homogeneous, except for being influenced only by environmental variation or random error.

(4) We changed the description of Figure 3b to explain that a particular color shows a range (not its boundary) of log2 fold change (FC) relative to the control.

(5) Corrected.

(6) Corrected.

Version 2

Abstract

Abstract

The loss of a single chromosome in a diploid organism halves the dosage of many genes and is usually accompanied by a substantial decrease in fitness. We asked whether this decrease simply reflects the joint damage caused by individual gene dosage deficiencies. We measured the fitness effects of single heterozygous gene deletions in yeast and combined them for each chromosome. This predicted a negative growth rate, i.e. lethality, for multiple monosomies. However, monosomic strains remained alive and grew as if much (often most) of the damage caused by single mutations had disappeared, revealing an exceptionally large and positive epistatic component of fitness. We looked for functional explanations by analyzing the transcriptomes. There was no evidence of increased (compensatory) gene expression on the monosomic chromosomes. Nor were there signs of the cellular stress response that would be expected if monosomy led to protein destabilization and thus cytotoxicity. Instead, all monosomic strains showed extensive upregulation of genes encoding ribosomal proteins, but in an indiscriminate manner that did not correspond to their altered dosage. This response did not restore the stoichiometry required for efficient biosynthesis, which probably became growth limiting, making all other mutation-induced metabolic defects much less important. In general, the modular structure of the cell leads to an effective fragmentation of the total mutational load. Defects outside the module(s) currently defining fitness lose at least some of their relevance, producing the epiphenomenon of positive interactions between individually negative effects.

Introduction

Cells and entire organisms can tolerate a surprisingly large number of deleterious mutations, especially when they affect gene function partially rather than completely, for example, when only one of two alleles is inactivated. In fact, a typical human germline cell has as many as about one hundred inactive genes, usually complemented by active alleles (MacArthur et al., 2012). Somatic cell lines in healthy tissues can add dozens of new (nearly always heterozygous) mutations every year (Moore et al., 2021). Understanding how this high burden of DNA damage translates into a relatively low mutational load, defined as a reduction in fitness, has proven to be a real challenge (Henn et al., 2015). One possible explanation is that mutations typically have negligible effects, especially when heterozygous. Another is that deleterious mutations are balanced by compensatory mutations, although the latter may not yet have been identified. However, it is also possible that the negative effects of mutations are significantly reduced when they are numerous, i.e. they effectively offset each other through positive epistasis.

Because the fitness architecture of complex organisms is particularly difficult to disentangle, researchers have turned to simpler systems. In the model organism Saccharomyces cerevisiae, collections of engineered single-gene mutations, such as complete deletions, have been created and used to estimate selection and dominance coefficients, as well as the effects of interactions between a few (typically two) loci (Costanzo et al., 2010; Giaever et al., 2002; Steinmetz et al., 2002). It would be much more difficult to combine large numbers of deleterious mutations. This inevitably long process would expose the strains under construction to a high probability of developing fitness-restoring genetic adaptations, which are known to occur even with limited cell proliferation (Szamecz et al., 2014). A rapid way to generate genomes with extensive loads of known mutations is needed. With this in mind, we noted that the budding yeast genome consists of approximately five thousand genes, distributed in a largely random manner across 16 chromosomes. We reasoned that monosomy, the loss of one chromosome in an otherwise regular diploid (2n-1), would mimic the sudden mutation (partial inactivation) of numerous and randomly assembled genes.

Results

Viability of monosomic strains

We began our study by inserting three markers into one chromosome within each homologous pair. These were two URA3 genes, each at the center of every arm, and one copy of kan near a centromere (Tutaj et al., 2022). Since its homolog was unmarked, mutants lacking the markers signaled a possible loss of the marked chromosome (Materials and Methods). We expected that such generated true monosomic strains (M1, M2, …, M16) would show visible growth defects, which we indeed observed, except for M1 (Fig. 1a). The loss of markers and slow growth were only hints, and often turned out to be false when DNA sequencing revealed that the expected loss of an entire single chromosome did not occur, or that it did occur but was accompanied by other changes. Fig. 1b shows isolates that passed the sequencing test. As shown in Fig. 1c, small colonies formed by monosomic cells were regularly invaded by much larger ones. For each monosomic, dozens of the latter were re-streaked and found to reproduce the fast-growing phenotype almost universally. To prevent the monosomics from overgrowing by the fast-growers, we had to keep the cultures small, down to 5 µl of YPD medium in the case of M5 or M15. The cultures had to be replicated (in tens or hundreds) so that a sufficient number of those dominated by the slow-growing cells could still be found and replenished.

Diploid yeast strains lacking single chromosomes (monosomics).

(a) Representative colonies of wild-type diploid (WT) and monosomic mutant strains (M1…M16) when grown together for 24 hours at 30 °C on agar-solidified YPD medium. (b) DNA whole genome sequencing coverage. (c) Examples of large colonies emerging among the small monosomic ones.

The procedure described above could not be completed for three chromosomes-VII, XII, and XIII-although we paid special attention to them. Regarding the screens for M7 and M13, we did not find a single mutant that lacked the markers, grew slowly, and could rapidly revert to a normal phenotype. Instead, we regularly encountered isolates of the desired genotype with no growth defect and a diploid genome. Hypothetically, two mitotic crossovers within a pair of homologous chromosomes could produce an offspring cell without the markers. We were able to reject this possibility based on the results of the present study and a previous study (Tutaj et al., 2022). Relevant data and calculations are provided in the Supplementary Text. The best explanation is that the marked chromosomes VII and XIII were lost and their unmarked homologs underwent endoreduplication shortly thereafter. The endoreduplication could result rom non-disjunction of a monosomic chromosome. Thus, we suggest that the cells of M7 and M13 were viable but suffered from very slow growth, high rate of endoreduplication, or both, and were thus especially rapidly overgrown by revertants. The search for M12 yielded unusually numerous and genotypically variable isolates, but none that were correct in terms of growth phenotype or DNA copy number mapping (Materials and Methods, Supplementary Text). Chromosome XII contains the only yeast rDNA region and is a strong recombinational hotspot (Tutaj et al., 2022). Therefore, we suspended attempts to derive M12, especially since, as we show below, its predicted fitness loss from haploinsufficiency was well within the range covered by other chromosomes.

In the wild, monosomic isolates have only been found for a few and the shortest chromosomes (Peter et al., 2018). Similar examples of monosomy have been detected in mutation-accumulation experiments (Sui et al., 2020; Zhu et al., 2014). It is uncertain whether these monosomies were simply tolerated, compensated by other changes, or conditionally adaptive in certain environments. There have been laboratory experiments suggesting that all monosomies can be induced in yeast, though indicated only by loss of heterozygosity at selected loci, but they may have been transient (Beach et al., 2017; Reid et al., 2008). We suggest that past work on yeast monosomy should be taken with a grain of salt and that future work should be properly planned. Colonies of putative monosomics may actually be diploids or aneuploids. Even if they were initially truly monosomic, they could easily have been overtaken by endoreduplications or other compensations for chromosome loss. This could happen even if a moderately sized colony was formed from an initially monosomic cell. Here, we combined the marker-based approach with genome-wide assays based on sequencing of DNA (reported above) and RNA (below) to confirm that nearly all yeast monosomics are viable.

Epistasis for fitness

Before studying epistasis between multiple mutations, their individual effects must be known. With respect to yeast growth rate, virtually none of the single gene deletions can increase it (Sliwa and Korona, 2005). About 1/5 of them stop growth altogether and another 1/5 slow it down detectably (Giaever et al., 2002). These estimates are based on homozygous or haploid strains under optimal laboratory conditions. Heterozygous deletions, again under good conditions, often show morphological changes, but their growth rate, i.e., proliferative fitness, is rarely and only modestly affected (Deutschbauer et al., 2005; Marek and Korona, 2016; Ohnuki and Ohya, 2018). In Materials and Methods, we explain how we selected 468 heterozygous single deletion strains as potentially (haplo)insufficient for growth on the rich medium, YPD. We performed repeated growth rate measurements for single deletion strains and compared them to a set of carefully selected control strains to make our estimates as accurate and reliable as possible. The doubling rates of individual heterozygous deletions were divided by that of the control to obtain relative doubling rates, rDRs (Datasheet 1) The frequency distribution of rDR is shown in Fig. 2a. It shows that the negative effects of the deletions were undoubtedly present, although predictably small, a substantial fraction of them were in the range of bidirectional effects, most likely composed of phenotypic plasticity and measurement error. (The existence of phenotypic variation not attributable to gene deletions is revealed by the variation within a genetically homogeneous control).

Contribution of positive epistasis to fitness of monosomics.

(a) Growth performance of heterozygous single-gene deletion strains tested in this study. Left: list of chromosomes with numbers of assayed deletion strains. Right: frequency distribution of rDR (doubling rate related to that of the control). (b) Growth performance of monosomics. Colored bars represent expected performance calculated as a sum of the single-gene effects per chromosome, rDRE=1+∑di, where di=rDRi−1 (deviation from the control). Gray bars show the observed performance of monosomic strains, rDRM. White arrows mark the expected departure from wild-type fitness (expected genetic load), gray the observed one. Black arrows show the extent and direction of epistasis. Three monosomics were not included in these assays (see the main text).

Epistasis is absent when an effect of the gene variant does not depend on the genetic content of other loci. In the case of fitness, a non-neutral mutation corresponds to a proportional change in wild-type fitness (Crow and Kimura, 1971). The absence of epistasis means that the fitness quotient of a given mutation remains unchanged regardless of whether it is the only one present in an individual or accompanied by other fitness-affecting mutations. Accepting this postulate is equivalent to adopting a multiplicative model of fitness structure, in which the fitness of a genotype involving n loci is the product of n respective quotients. The model is additive for log-fitness because such a transformation turns a quotient into a deviation from one and a product of quotients into a sum of such deviations. Since we are working with a (log2) transformed measure of fitness, the deviation caused by an i-th gene deletion is equal to di = rDRi−1. The expected effect of multiple loci is the sum of all deviations involved, rDRE=1+∑di. To predict how each of the 16 monosomic strains should perform, we summed the deviations caused by the individual deletions present on each chromosome (obtaining 16 values of rDRE). We summed both negative and positive values of d to account for non-genetic variation of estimates. We then compared the predicted values of rDRE with the corresponding values of rDRM, that is, estimates obtained experimentally for the actual monosomic strains (Datasheet 1). Fig. 2b shows the expected and obtained values of monosomics’ rDR, where epistasis is equal to the difference between them. The epistasis between multiple deleterious gene deletions turned out to be not only positive but also large, much larger than that observed for just pairs of deletions (Jasnos and Korona, 2007). Most strikingly, some monosomic strains were expected to have a negative doubling rate, i.e. were predicted to be effectively lethal, but turned out to be able to proliferate. For several monosomics, and especially M4, most of the predicted mutational load was canceled out by epistasis. The contribution of epistasis to fitness seemed so large that it might be difficult to accept without a functional rationalization.

Transcriptome reaction

Epistasis can be considered in a purely abstract way, as a deviation from additivity/multiplicity. But it must have a biological explanation. Our attempt to find it began with the isolation and quantitative analysis of mRNA from the ancestral diploid strain and eight monosomic strains. (The selection was random. It nevertheless resulted in the inclusion of strains in which the level of epistasis was low, M1 and M8, or high, M2 and M16.) The first question was obvious: was the decrease in gene dosage compensated by increased expression? Fig. 3a shows that there was no detectable increase in the average intensity of transcription on the monosomic chromosomes. These averages depend on hundreds of genes and would therefore be largely insensitive to increased expression of important but relatively few genes. In this experiment, we knew which genes would be most rewarding to upregulate, those that were most haploinsufficient, and could focus our attention on them. Fig. 3b shows that the fitness effect of a single gene-deletion did not correlate with the expression of this gene in a monosomic strain. Thus, neither physical underrepresentation (single copy) nor functional importance (impact on growth) triggered an increase in expression of the genes on monosomic chromosomes: their mRNA level was halved on average although individual genes could show either up- or downshifts.

Absence of transcriptional compensation in monosomic strains.

(a) Halved RNA production on monosomic chromosomes. For every ORF, the obtained number of RNA-seq reads was divided by the number expected for it under expression being constant over an entire genome. (b) Expression under monosomy vs. single-deletion fitness (rDR). X-axis shows the length of a monosomic chromosome with centromeres marked as circles and gene deletions as bars; colors show the effect of monosomy on the level of a particular mRNA with a particular color showing a range of log2 fold change (FC) relative to the control. Y-axis: the difference in rDR between a single gene deletion strain and the control, d=rDR-1. The correlation between fitness effect (rDR-1) and shift in expression (log2FC) is reported as Spearman’s coefficient rs with associated and p-value (not corrected for the multiplicity of comparisons).

Monosomy could have altered the expression of any genes, not just those on the affected chromosomes. We aimed to find any distinct and functionally interpretable patterns of adjustment that might explain how the functioning of monosomic strains was perturbed. We found multiple statistically significant up- and downregulations in the transcriptomes of each monosomic strain analyzed (Datasheet 2). Both parallelisms and incongruencies between them were visible.

Fig. 4a shows the similarities. The translational apparatus of the cytosol was on average significantly and universally upregulated, as evidenced by transcripts encoding proteins that build both large (e.g. RPL28) and small (RPS29B) ribosomal subunits. Cytosolic proteolysis was downregulated (core proteasome component PRE6). In remarkable agreement, chaperones required to fold newly synthesized peptides were upregulated (SSB1/2), while those required to direct destabilized chains to degradation were downregulated (SSA1, HSP82, HSP104). Transcripts encoding mitochondrial proteins were generally less abundant, with a marked decrease in the expression of several genes encoding the electron transport chain (Datasheet 2). In parallel, the expression of the antioxidant machinery was downregulated, especially that of a major ROS scavenger (SOD2).

Parallel and divergent shifts in transcriptomes of monosomic strains.

Heat maps show monosomic mRNA frequencies divided by respective diploid (control) ones. (a) Gene Ontology categories selected to demonstrate similarities in transcriptional profiles of monosomic strains. (b) Regulons demonstrating differences in gene expression between monosomic strains. Expanded versions of all panels can be found in Supplementary Figure S1.

To find transcriptomic differences, we examined groups of genes, each responding to a known signal. This increased the chances of detecting statistical significance and functional divergence. In our experiment, all strains received identical external signals. Therefore, different expression patterns would indicate different internal perturbations. Indeed, as shown in Fig. 4b, substantial variation was detected within several regulons: cAMP-PKA (glucose-activated signaling), NCR (nitrogen catabolism repression), GAAC (general amino acid control), and Msn2up (activation by a broad range of stresses). These differences underscore the significance of the above reported uniformity in the pattern of biosynthetic upregulation and proteolytic downregulation.

All of the above considerations refer to the relative abundance of mRNA species. We also attempted to compare the absolute size of the transcriptome in wild-type BY and three monosomic strains: M1, M2, and M3. We added an admixture of specific external mRNA to provide a quantitative reference, or “spike”, to known total cell volumes of each strain and repeated our assays. The results are shown in Supplemental Figure S3. In summary, the spike represented 4.04% of total mRNA for the BY sample and 6.08, 15.6, and 24.2% for the respective monosomic strains. Thus, monosomy would be associated with a decrease in the absolute level of mRNA. A possible caveat is that the monosomic cells had an altered morphology. They were more rounded in shape and approximately 60% larger in long axis, so that their individual volumes were several times larger than wild-type. Thus, the size of the transcriptome for M1 and M2 would not be reduced but actually increased if calculated per whole cell (one nucleus). Indeed, one explanation for the observed epistasis for viability could be an ample overproduction of all transcripts, so that even those halved by monosomy are sufficiently abundant. We consider the cell volume calculations more appropriate and therefore accept that the transcripts from monosomic chromosomes were even less abundant than half of the wild strain abundances. The apparent decrease in the cellular density of the transcriptome in the cytoplasm of monosomics is not paradoxical because the RNA content, including ribosomal protein mRNAs, decreases with slowing growth (García-Martínez et al., 2016; Warner, 1999). However, even if one accepts that monosomic transcriptomes were expanded (per nucleus), they would still be unbalanced. Thus, if the lack of stoichiometry between proteins and the resulting low efficiency of ribosome assembly, rather than the simple scarcity of specific elements, is the major consequence of monosomy, it would remain so regardless of the absolute size of the transcriptome.

Discussion

The successful derivation and maintenance of monosomic strains demonstrated that it is possible to obtain an a priori designed set of strains with hundreds of partially inactive genes, dozens of which are known to have deleterious fitness effects. It implies that the ability to remain viable despite carrying extensive mutational burdens is not limited to specific combinations of genes, but would likely apply to a wide variety of them. Viability was supported by the epistatic effect, which was of positive value and remarkable strength, but may still be underestimated in our study. It was shown that many of the single deletions with negative fitness effects had undergone compensatory evolution under recurrent replication needed to maintain the strains carrying them (Puddu et al., 2019). Since we used the same collection of deletions, it is likely that some of them were at least partially compensated. All of these effects that may have been missing in the individual deletions were present in the newly derived monosomic strains. This caveat actually strengthens our claim that the positive epistatic component can be substantial and even offset most of the total damage associated with the combined individual negative effects.

The advent of systems biology has raised hopes that the “statistical” and “biological” sides of epistasis can be coherently brought together (Moore and Williams, 2005; Phillips, 2008). Indeed, a truly functional explanation of the measured growth effects would require an understanding of how the material and energetic expenditures were altered by the same mutations when they were separate or combined. Such a metabolomic interpretation could only be attempted semi-quantitatively and for much simpler systems (Molenaar et al., 2009). Regarding proteome and transcriptome, the latter has the advantage of higher accuracy and repeatability, the ability to detect the products of even weakly expressed genes and, crucially for this study, to provide clear signals of major functional transitions such as fast-slow growth, anabolism-catabolism, presence-absence of stress. The relationship between the transcriptome and the proteome is not straightforward in multicellular organisms, where compensation for the perturbations introduced by aneuploidy is often observed (Birchler and Veitia, 2021). In yeast, most of the changes observed in mRNA are generally reflected as corresponding shifts in both the profile of translation and the composition of mature proteins (Larrimore et al., 2020; Pavelka et al., 2010). The relationship is not perfect as, for example, disomics have a somewhat attenuated average expression of proteins encoded on the duplicated chromosome. But even there, the result is still much closer to doubling than to parity (Dephoure et al., 2014). Post-transcriptional attenuation can be more frequent and pronounced when only single genes are doubled, but this is a very different case from ours (Ascencio et al., 2021). Crucially, while signs of compensation at the proteome level are found in wild aneuploids, they are much less pronounced in the laboratory strains and especially in the BY used here (Messner et al., 2023; Muenzner et al., 2024). (It is tempting to suggest that the wild aneuploids represent a biased sample of aneuploid mutants, i.e., they either already possessed means of compensating for the distortions introduced by aneuploidy or were able to evolve them and thus avoid purging by selection.) In summary, although mRNA analysis does not provide a complete description of the metabolism of the laboratory-generated monosomic cells, it does provide valuable information about it.

The simultaneous upregulation of genes encoding ribosomal proteins (RP) and downregulation of those encoding subunits of the proteasome was observed in all monosomic strains, providing a crucial insight into the functional response of the cell to monosomy. The absence of a single chromosome meant that several genes encoding the translational machinery that resided on it were halved in dosage, while dozens of others remained unaffected. The cell could not restore the required stoichiometry by overexpressing the affected genes. Favorable environmental conditions signaled that translation needed to be intensified, but functioning ribosomes were in short supply, resulting in seemingly indiscriminate overproduction of ribosomal proteins and withholding of degradation of both them and other cytosolic proteins. The response was inadequate and costly, but it should be seen as an attempt at ad hoc rebalancing rather than a prepared (evolved) response. This interpretation, strongly suggested by the transcriptomic data, appears even more plausible when analyzing the functional diversity of the 468 genes qualified as haploinsufficient. Table 1 shows that the diversity was ample: of the one hundred Slim GO Biological Process categories, dozens could be linked to each chromosome-associated subset of deletions. However, there was a significant common motif. Processes involved in the “protein synthesis apparatus” (our term) were most frequently represented. Their predicted damage was also particularly high, typically sufficient or exceeding that required to reduce the growth rate to the levels actually observed in monosomics. The biosynthetic perturbation was thus real, and its relationship to the observed transcriptomic response was likely causal, not merely coincidental.

Some of the other adaptations seen in the monosomal transcriptomes could also be related to the dominant effect of translational inefficiency. A signal to increase biosynthesis would mobilize fermentation and reduce oxidation, which would downregulate genes encoding ROS-scavenging proteins. The Mns2-mediated and other stress responses were apparently absent, contrary to previous reports (Sheltzer et al., 2012). The hallmarks of stress responses include decreased RP production and increased proteolysis (to remove destabilized proteins). Here, these two processes were driven in opposite directions, blurring the expected patterns. The non-homogeneity of the response in other regulons analyzed - cAMP-PKA, NCR, GAAC - means that functions other than ribosome assembly were severely affected, but differently, depending on the composition of genes made insufficient by the loss of a chromosome (see also Datasheet 2). If there were so many different piecemeal responses, why was the uniform positive epistasis observed?

A possible answer begins with the assumption that the cell is an aggregate of multiple functional modules (Hartwell et al., 1999). Our monosomic strains had many different modules that were negatively affected by partially inactivating mutations. These mutations did not interact with each other directly, but rather through the modules to which they belonged. Mutations that affected the most critical module(s) suppressed the negative effect of mutations that affected other, non-limiting modules. The latter were either less damaged or less needed under current conditions. The epiphenomenon of positive epistasis for fitness reflected the fact that not all mutations exerted their negative effects, at least not to the full extent. In our case, this is not just speculation, but rather the simplest and most prudent explanation linking growth rate and transcriptomic data. Our experimental system is special in a way: a microbial cell in which several modules have been compromised but exposed to favorable conditions and thus tested for the single capacity to grow rapidly. Although peculiar, the system is also instructive. Any unicellular organism is typically tested for only a few specific condition-dependent functions defining fitness, which are very different under, e.g., growth or starvation. Similarly, the specialized cells of a complex organism are required to support only a few, and different, elements of life. The modularity of the cell helps to understand how so many partial genetic defects can be carried by so many different organisms without drastically reducing their fitness: rarely or never are all the defects really important.

Interactions within pairs of gene knockouts have received much attention (Costanzo et al., 2020; Segrè et al., 2005; Szappanos et al., 2011). Epistasis for fitness arising from interactions between at least several mutations has also been studied but mostly in the context of “diminished returns” of successive beneficial variants (Barrick et al., 2009; Chou et al., 2011; Khan et al., 2011; Kryazhimskiy et al., 2014). It has been proposed that the pattern may arise from interactions of mutations through a ‘global’ phenotypic interface, such as fitness (MacLean et al., 2010; Perfeito et al., 2014; Schoustra et al., 2016). However, a similar overall effect could also be produced by “idiosyncratic” interactions without any mediating factors (Lyons et al., 2020; Reddy and Desai, 2021). Indeed, this has been demonstrated experimentally, at least for a selected and moderately large set of mutations (Bakerlee et al., 2022). Finally, negative epistasis of beneficial mutations has been attributed to cellular modularity by postulating that the positive contribution of each functional module to fitness must have its upper bound (Wei and Zhang, 2019). Given that the negative epistasis between multiple positive effects can arise from different mechanisms, we do not insist that the explanation of positive epistasis between multiple negative effects we propose is the only possible one. However, it may be particularly applicable when the deleterious mutations are truly numerous and distributed across many cellular subsystems that work as functional modules.

Materials and Methods

Single gene deletion strains

Selection of single gene deletions with a possible effect of haploinsufficiency

Deutschbauer et al. have assayed a complete collection of heterozygous single gene deletion strains and identified a total of 184 genes, 98 essential and 86 non-essential, as haploinsufficient for growth in rich medium, YPD (Deutschbauer et al., 2005). We included this set of genes in the present study. Using a different technique, Marek et al. have tested al1 142 essential heterozygous single-gene deletions and 946 non-essential deletions selected as likely not neutral for growth (Marek and Korona, 2016). We reviewed the latter study and, using a false discovery rate of 0.15, accepted up to 404 genes, 256 essential and 148 non-essential, as potentially haploinsufficient. The two sets obtained in two different studies, 184 and 404, overlapped in 112 cases. The overlap was much higher than expected, 14.3, if the two sets were just random samples from among 5,200 yeast genes. On the other hand, it was limited and suggested that new growth assays were desirable. The new assays are described below; they included all the unique strains identified in the two studies minus 5 that were not present in our current strain collections and another 3 that were dropped as superfluous for the orthogonal design of experimental blocks described below. (They were also the 3 least promising based on the earlier assays.) Datasheet 1 lists the strains.

Control for single deletion strains

To correctly quantify the predictably small negative growth effects introduced by heterozygous single deletion strains, an unbiased and accurate estimate of a wild-type phenotype is required. In the present study, we attempted to achieve this by using multiple strains as controls rather than a single strain. The reason for this was that the gene deletions used here were constructed over several years by several laboratories and then treated with repeated rounds of propagation, which could lead to genetic divergence. We felt it was risky to use one strain as a control for all the others. We looked for a group of strains that were indistinguishable in terms of growth rate, suggesting that they had not acquired genetic changes that affected this trait. In the case of non-essential genes, we searched the Saccharomyces Genome Database and found 25 ORFs originally labeled “dubious”, which are currently almost certainly spurious ORFs, located between other ORFs and no closer than 100 bp to their START or STOP codons. Heterozygous strains carrying deletions of these genes were tested for doubling rate in a manner specific to this study (described below). After repeated assays, 16 deletion strains that were closest to the medium growth performance and not statistically different from each other were selected as controls. Knowing that the non-essential and essential strains differ somewhat in their origin and subsequent handling (Brachmann et al., 1998), we decided to derive a separate set of control strains for the latter. We selected 32 essential genes that were in the very center of a single, strong and narrow modal peak of the frequency distribution of estimates collected for the entire collection of heterozygous essentials (Marek and Korona, 2016). Again, after repeated tests of growth rate, 16 strains that were closest to the median and not different from each other were selected to serve as the final control for essential gene deletions. Control strains from both groups are listed in Datasheet 1.

Monosomic strains

Parents of monosomic strains

In an earlier study, we used 32 strains that had a counter-selectable marker (URA3) near the center of each of the 32 chromosome arms and a drug resistance marker (kan) near a centromere (Tutaj et al., 2022). Pairs of strains with the same centromere marker and the counter-selectable markers located on either the left or right arm of the same chromosome were mated with each other. The resulting diploids were sporulated and tetrads dissected to get triple-marked haploid genotypes, URA3-kan-URA3. The latter were mated with a standard BY haploid strain of the opposite mating type. In the final set of 16 diploid strains, each strain had one chromosome triple-marked while its homolog and the remaining 15 chromosome pairs were isogenic with the diploid strain BY4743.

Derivation and verification of monosomic strains

The diploid strains with triple-marked chromosomes were grown overnight in synthetic complete medium (SC) and then 50 to 500 µl samples of the resulting cultures were plated on standard 5-FOA (5-fluoroorotic acid) plates. Emerging colonies were transferred to new 5-FOA plates and YPD plates with 200 mg/ml geneticin. The goal was to identify variants that were able to grow on the 5-FOA plates but not on the geneticin plates, indicating that all three marker genes may have been lost along with an entire chromosome. Colonies identified in this way were often of different sizes, suggesting that they were genetically heterogeneous. The next criterion was to find variants that produced colonies that were visibly smaller than those of the parental strain, but rarely, though regularly, produced colonies similar to those of the parental strain. Such strains were grown in replicate small cultures (5 to 100 µl, depending on the monosomic strain), tested for negligible frequency of cells forming large colonies, and collected in larger samples that allowed isolation of DNA in quantities sufficient for next-generation sequencing. Monosomy was considered confirmed when the number of reads for the entire length of a single and expected chromosome was halved. Although simple, this protocol required multiple attempts for some chromosomes because few colonies tended to appear on the 5-FOA plates, most of them remained resistant to geneticin, and those that passed these two criteria did not show the required reversibility, that is, the tendency to occasionally return to normal growth in an apparent step. Even when all these phenotypic criteria were met, sequencing occasionally revealed genomes other than those of pure monosomic origin, that is, with the whole and only one chromosome removed. Nevertheless, once confirmed, the monosomic strains could have been reliably propagated on rich and synthetic media, including that containing 5-FOA, as long as the recurrent appearance of fast-growing colonies was monitored and counteracted.

Control for monosomic strains

The monosomic strains were all derived by us and the derivation involved our stock of haploid strains BY4741 and BY4742. These two were then crossed to produce a diploid BY4743 that lacked the URA3MX4 and kanMX4 cassettes and was used as a control in the growth assays of the monosomic strains.

Estimation of DR and rDR

The collection of single gene deletion strains was arrayed on 6 flat-bottomed 8×12 well titration plates with 150 µl aliquots per well. Within each plate, the first and last wells contained clean YPD medium, rows 3 and 10 contained control strains, and the deletion strains occupied the remainder of the plate. Plates were filled with either essential or non-essential deletions accompanied by the 16 control strains listed above. One plate contained both essential and non-essential deletions along with 8 essential and 8 non-essential control strains. Plates were inoculated at 1-5% from thawed samples and kept non-agitated at 30°C for 48 hours until they reached approximately similar densities of stationary phase cells. Such conditioned microcultures were used to inoculate plates with fresh YPD at 0.5% and maintained at 30°C with shaking at 1,000 rpm. Cultures were analyzed for OD (600 nm) every 0.5 hours using a TECAN Infinity reader. Four independent replicates of measurements were performed, starting with independent conditionings. OD readings were used to calculate DR (doubling rate). To obtain the rDR (relative doubling rate), each DR estimate of an experimental strain was divided by the average DR of the control strains present on the same plate. In the case of the monosomic strains, the entire protocol was analogous except that all experimental and control strains were kept in one plate. Cultures of monosomic strains used in this assay were tested to contain less than 1% fast-growing cells at the time of OD measurements.

Analysis of DNA and RNA

The first step in preparation for both DNA and RNA analysis was to collect samples of monosomic cells that would be nearly free of the rapidly growing revertant cells. Individual monosomic strains were grown as replicate microcultures (5 to 100 µl) in YPD at 30°C to stationary phase. The latter were serially diluted and plated to test for the appearance of large colonies indicating the appearance of compensatory mutations. The microcultures in which more than 99% of the colonies were typical of a particular monosomic strain were pooled. Such tested stationary cultures of monosomic cells were directly used to extract DNA as template for high coverage sequencing (PE 150, expected read depth ~80). The resulting reads were mapped along standard yeast chromosome sequences using bowtie2 (Langmead and Salzberg, 2012), along standard sequences of yeast chromosomes: Ensembl release 100, S. cerevisiae genome R64-1-1. Duplicate reads were marked with MarkDuplicates (Picard Toolkit 2019. Broad Institute, GitHub). Samtools (1.15.1) was used for BAM files sorting, indexing and coverage analysis (Danecek et al., 2021).

For RNA, purity-tested cultures of monosomic cells were transferred to fresh YPD and incubated for 4 hours at 30°C with agitation. Total RNA was then extracted using the RiboPure™ RNA Purification Kit. Three replicates of the monosomic and control BY4743 strain were prepared in this manner. Library preparation and PE 150 sequencing were performed by Novagene. Approximately 20 million read pairs were generated per sample. Quality control of the reads was performed using fastQC v0.11.9 (Andrew, 2010). RNA reads were aligned to the above standard sequence using Hisat2 v2.1.0 (Kim et al., 2015). The resulting alignment files were sorted and indexed with samtools (1.9). Transcript quantification was performed using cuffquant/cuffnorm v2.2.1 (Trapnell et al., 2012). Gene count data normalization (“TMM” method) and differential expression analysis (exact test) were performed in the EdgeR test (Robinson et al., 2010).

For transcriptome analysis with the addition of the mRNA spike, cultures were conditioned as above. The density of the cultures was estimated by counting the cells in the Burker’s chamber. The cells were also photographed under a light microscope and the long and short axes of the cells were measured for about 50 randomly selected cells. Using the formula for ellipsoids and assuming that the two shorter axes are equal, the volumes of individual cells and their sums were calculated. Equal amounts of the mRNA spike (ERCC RNA Spike-In Mix, TermoFisher) were added to all cell samples from which RNA was extracted using hot formamide (Shedlovskiy et al., 2017). Subsequent library preparation, sequencing and analysis were performed as described above.